---
title: "tmt pca analysis"
author: "Christine Berryhill"
date: "2022-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#open up files
```{r}

library(rstudioapi)
library(stringr)
library(ggplot2)
library(dplyr)
library(ggforce)

#Opening up all the files

temp = list.files(pattern="*.csv")
list2env(
  lapply(setNames(temp, make.names(gsub("*.csv$", "", temp))), 
         read.csv), envir = .GlobalEnv)

tmt_methylsites<-as.data.frame(
 apply(tmt_methylsites,2, function(x) gsub("\\s+", "", x)))

tmt_methylsites<- na.omit(tmt_methylsites)


```
# PCA analysis for both all peptides and just methylated peptides

#prepare data frame to do a pca analysis
  columns are peptides/Format 3
  rows are samples

```{r}
#Unenriched

tmt_mat<- tmt_methylsites[,c(19:28,40)]

tmt_mat<- tmt_mat[!duplicated(tmt_mat$Format3),]

rownames(tmt_mat)<- tmt_mat[,11]

tmt_mat<- tmt_mat[,-c(11)]

colnames(tmt_mat)<-c("DU4475_R1", "DU4475_R2", "DU4475_R3", "DU4475_R4","DU4475_R5",
                    "MCF10A_R1", "MCF10A_R2", "MCF10A_R3", "MCF10A_R4","MCF10A_R5")

cols.num<- colnames(tmt_mat)

tmt_mat[cols.num]<- sapply(tmt_mat[cols.num], as.numeric)

tmt_mat<- na.omit(tmt_mat)

tmt_mat_t<- t(tmt_mat)

```
```{r}

tmt_pca<- prcomp(tmt_mat_t)

```


Plotting the pca
```{r}

plot(tmt_pca$x[,1], tmt_pca$x[,2])

```
making a scree plot
```{r}
tmt.pca.var<- tmt_pca$sdev^2

tmt.pca.var.per<- round(tmt.pca.var/sum(tmt.pca.var)*100,1)

```
```{r}
barplot(tmt.pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

```

Using ggplot to make a nice pca plot
```{r}
tmt_pca_data<- data.frame(sample= rownames(tmt_pca$x),
                         x=tmt_pca$x[,1],
                         y = tmt_pca$x[,2])

tmt_pca_data$cell<-c(rep("DU4475",5) , rep("MCF10A",5))
```
```{r}

a<-ggplot(data=tmt_pca_data,aes(label=sample,x=x,y=y, color = cell)) + 
 
  ggtitle("PCA Plot of Kme sites in Dataset Without Enrichment") +

  geom_point(aes( size = 4)) +
  
  scale_color_manual(values = c("DU4475"= "#88a4bf", "MCF10A"="#c59b19"))+

  xlab(paste("PC1: ",tmt.pca.var.per[1],"%",sep="")) +

  ylab(paste("PC2: ",tmt.pca.var.per[2],"%",sep=""))+
   #geom_mark_ellipse(aes(color = as.factor(cell)), expand = unit(0.5, "mm"), show.legend = F )+
  #geom_encircle(aes(group = type), color = "red", s_shape = 0.01, expand = 0.04)+
  theme_classic()+
  theme(legend.text = element_text(size = 10))+
  guides(color=guide_legend(override.aes = list(size = 5)),
         shape = guide_legend(override.aes = list(size=5)))
 
a

ggsave("TMT.methyl.pcaplot.eps", a, device= "eps", width = 6, height = 3)
```

looking at what contributes the most to PC1
```{r}

#getting the loading scores

loadingScores<- tmt_pca$rotation[,1]

#getting the magnitudes of loading scores

geneScores<- abs(loadingScores)

#soring from high to low

geneScoreRanked<- sort(geneScores, decreasing = T)

#getting top 5 format 3

top5genes<- names(geneScoreRanked[1:10])

top_tmt_pca<-tmt_pca$rotation[top5genes,1]
top_tmt_pca

```

# PCA Analysis of the proteomes using the normalized protein amounts
```{r}
#PCA Analysis of proteins

tmt_mat_protein<- tmt_prot[,c(17:26,4)]

tmt_mat_protein<- tmt_mat_protein[!duplicated(tmt_mat_protein$Accession),]

rownames(tmt_mat_protein)<- tmt_mat_protein[,11]

tmt_mat_protein<- tmt_mat_protein[,-c(11)]

colnames(tmt_mat_protein)<-c("DU4475_R1", "DU4475_R2", "DU4475_R3", "DU4475_R4","DU4475_R5",
                    "MCF10A_R1", "MCF10A_R2", "MCF10A_R3", "MCF10A_R4","MCF10A_R5")


tmt_mat_protein<- na.omit(tmt_mat_protein)

tmt_mat_protein_t<- t(tmt_mat_protein)

```
```{r}

tmt_pca_protein<- prcomp(tmt_mat_protein_t)

```

Plotting the pca
```{r}

plot(tmt_pca_protein$x[,1], tmt_pca_protein$x[,2])

```
making a scree plot
```{r}
un.protein.pca.var<- tmt_pca_protein$sdev^2

un.protein.pca.var.per<- round(un.protein.pca.var/sum(un.protein.pca.var)*100,1)
```
```{r}

barplot(un.protein.pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

```

Using ggplot to make a nice pca plot
```{r}
tmt_pca_protein_data<- data.frame(sample= rownames(tmt_pca_protein$x),
                         x=tmt_pca_protein$x[,1],
                         y = tmt_pca_protein$x[,2])

tmt_pca_protein_data$cell<-c(rep("DU4475",5) , rep("MCF10A",5))

```
```{r}

e<-ggplot(data=tmt_pca_protein_data,aes(label=sample,x=x,y=y, color = cell)) + 
 
  ggtitle("PCA Plot of Proteins in Dataset Without Enrichment") +

  geom_point(aes(color = cell, size = 4)) +
  
  scale_color_manual(values = c("DU4475"= "#88a4bf", "MCF10A"="#c59b19"))+

  xlab(paste("PC1: ",un.protein.pca.var.per[1],"%",sep="")) +

  ylab(paste("PC2: ",un.protein.pca.var.per[2],"%",sep=""))+
  
  theme_classic()+
  theme(legend.text = element_text(size = 10))+
  guides(color=guide_legend(override.aes = list(size = 5)),
         shape = guide_legend(override.aes = list(size=5)))
  #geom_encircle(aes(group = cell), color = "black", s_shape = 0.01, expand = 0.04)
e
ggsave("TMT.protein.pcaplot.eps", e, device= "eps", width = 6, height = 3)
```


looking at what contributes the most to PC1
```{r}

#getting the loading scores

loadingScores<- tmt_pca_protein$rotation[,1]

#getting the magnitudes of loading scores

geneScores<- abs(loadingScores)

#soring from high to low

geneScoreRanked<- sort(geneScores, decreasing = T)

#getting top 5 format 3

top5genes<- names(geneScoreRanked[1:10])

top_tmt_pca_protein<-tmt_pca_protein$rotation[top5genes,1]
top_tmt_pca_protein

```
